---
output:
  html_document:
    keep_md: yes
---
# Reproducible Research: Peer Assessment 1
First, I load necessary libraries with messages supressed.  Doing this once so future messages aren't silently supressed
```{r, message=FALSE}
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data

First I load the data and transform the date column into a more usable format

```{r}
activity <- read.csv("activity.csv", header=TRUE)
#Convert the dates
Date <- strptime(activity$date, "%Y-%m-%d")

#Create a logical vectors of everything other than the original date
#then bind those columns with the new Date back to our data frame
lv <- names(activity) != "date"
activity <- cbind(activity[,lv], Date)

#Does everything look ok?
summary(activity)
```

## What is mean total number of steps taken per day?
```{r}
totalByDay = activity %>%
              group_by(Date) %>%
              summarise(sum=sum(steps))

hist(totalByDay$sum, breaks=10, main = "Histogram of total daily steps",
     xlab = NULL)
```

Calculate mean and median of the daily totals, ignoring days that totalled to NA values. Based on the histogram I am expecting values between 10,000 and 11,000

```{r}
mean(totalByDay$sum, na.rm=TRUE)
median(totalByDay$sum, na.rm=TRUE)
```

## What is the average daily activity pattern?

```{r}
averageByInterval <- activity %>%
                      group_by(interval) %>%
                      summarise(avg=mean(steps, na.rm = TRUE))
plot(averageByInterval, type="l", ylab ="Average number of steps")
```

Which interval has the maximum average number of steps?
```{r}
averageByInterval <- activity %>%
                      group_by(interval) %>%
                      summarise(avg=mean(steps, na.rm = TRUE))

averageByInterval[averageByInterval$avg==max(averageByInterval$avg),]
```

## Imputing missing values
How many missing values are we dealing with?  Is it a large proportion?
```{r}
sum(is.na(activity$steps))
sum(is.na(activity$steps)) / nrow(activity)
```
I will replace the missing values with the average number of steps taken during that interval for which we have values

```{r}
# Get just the rows with missing steps
missing <- activity[is.na(activity$steps),]

# Add a column for imputed values. This combination of mutate()
# and sapply() is a little messy -- there is probably a nicer way
missing <- mutate(missing, 
                  steps=sapply(interval, 
                               function(y) averageByInterval[averageByInterval$interval == y,]$avg))

# Create a new data frame with 1) original rows with non-null steps 2) newly created data points

newActivity = rbind(activity[!is.na(activity$steps),],
                    missing)
```
Lets check that we have the same number of rows we started with and that there are no null values in steps any more
```{r}
nrow(activity) == nrow(newActivity)
sum(is.na(newActivity$steps))
```

Now I recreate our original histogram and recacluate the mean and median with our new data set.
```{r}
newTotalByDay = newActivity %>%
                group_by(Date) %>%
                summarise(sum=sum(steps))

hist(newTotalByDay$sum, breaks=10, 
     main = "Histogram of total daily steps",
     xlab = NULL)
mean(newTotalByDay$sum)
median(newTotalByDay$sum)
```
Imputing missing values has **virtually no impact** on my estimates of total number of steps taken each day.
```{r}
format(mean(newTotalByDay$sum) / mean(totalByDay$sum, na.rm = TRUE), nsmall=5)
format(median(newTotalByDay$sum) / median(totalByDay$sum, na.rm = TRUE),nsmall=5)
```
## Are there differences in activity patterns between weekdays and weekends?
First I add a new factor variable indicating weekday or weekend
```{r}
newActivity <- newActivity %>%
               mutate(dayType = "weekday")
newActivity[weekdays(newActivity$Date) %in% c("Saturday", "Sunday"),"dayType"] <- "weekend"
newActivity$dayType <- as.factor(newActivity$dayType)

newAverageByInterval <- newActivity %>%
                        group_by(interval, dayType) %>%
                        summarise(avg=mean(steps))

```
Is the average by interval different for weekdays and weekend?  In general yes:  
  
- Weekday has a higher peak in the morning  
- Weekends are a little more dispersed during the day
```{r}
ggplot(newAverageByInterval) + 
  geom_point(aes(x=interval,y=avg)) + 
  facet_wrap(~dayType)
```